---
title: "CashflowModelBasics"
author: "DavidSalisbury"
date: "13 June 2016"
output: html_document
---

In an actuarial model *contracts* can be modelled as a collection of *cashflows*.

A cashflow has three key properties
1. Time it occurs
2. The amount of the cashflow
3. The probability it occurs

There is a fourth property usually included as well
4. A discounting function to give the present value of the cashflow.

So a cashflow can be represented as 4-tuple: $(t,C,p,v)$

And a contract can be represented as a 3-tuple of functions of time $t$, $(C(t), p(t), v(t))$.

These functions should be welldefined for any integer t >=0.

Let's start with a very simple and well known example... An annuity certain payable for 10 years in advance, with discount rate = 0/no discounting.

Here 
``` {r}
C <- function(t) {ifelse (t < 10, 1, 0)}
p <- function(t) {1}
v <- function(t) {1}
CF <- function(t) {(p(t)) * C(t)}
DCF <- function(t) {v(t) * p(t) * C(t)}
CF(5)
DCF(10)
PV <- function(n) {Reduce("+",Map(DCF,seq(0,n)),0)}
PV(100)
```


## Adding discounting

So far so simple. Now let's tweak the discount function to get a discounted annuity certain...

In order to do it we need a handy little utility function that can repeat the last element of a vector as many times as we need something in order to make a vector of arbitrary length $n$.

``` {r}
repeatLast <- function (v, n) 
  {
      c(head(v, -1), rep(tail(v,1) ,n))[1:n]
  }
```

Let's pick a single discount rate to start with, e.g. $2.8%$. 

```{r}
i <- 0.028
is <- repeatLast((i),120)
```

Now we can use the absolutetly brilliant function *Reduce* to get the cumulative product of $1/(1+i)$, using accumulate=TRUE gives us a vector of the cumulative results.

```{r}
v_generic <- function (t, i) {Reduce("*", 1/(1+(repeatLast(i,t+1))),1,accumulate = TRUE)[t+1]}
v <- function(t) {v_generic(t,(0.028))}
```


So now if we repeat the calculation...

``` {r}
CF <- function(t) {(p(t)) * C(t)}
DCF <- function(t) {v(t) * p(t) * C(t)}
CF(5)
DCF(5)
DCF(10)
PV <- function(n) {Reduce("+",Map(DCF,seq(0,n)),0)}
PV(100)
```

The last result box should give us the value of an annuity certain due payable for 10 years.

Let's check this with the traditional formula

```{r}
a_n_due <- function(i,t) {(1-(1+i)^-t)/(1-(1/(1+i)))}
a_n_due(0.028,10)
```

## Adding contingency

Next we want to add in the probability of payment.

As this is an actuarial calculation let's be old school and use a well loved familiar mortality table - the original 92 projection of pension male annuitant ammounts weighted mortality, arbitrarily chooding calendar year 2010 (PMA92C=2010)



```{r}
#TODO: Make this a relatie reference so we can bundle the file in a github repo
PMA92C10 <- read.csv("C:\\Users\\David\\Dropbox\\Code\\Data\\stripped pma92c10.xls - PMA92C10.csv")
```

First example let's take an annuity paid immediately (for ten years as before) to a surviving life aged at 40 at the start.

The probabilities need to be converted from qxs to cumulative pxs.

```{r}
qxs <- PMA92C10[PMA92C10$age >= 40,]$qx
pxs <- Reduce("*",1-qxs,1,accumulate=TRUE)
p <- function(t){repeatLast(pxs,120)[t+1]}

```


``` {r}
CF(5)
DCF(5)
DCF(10)
PV(100)
```

Already we can see that this implementation of p using a lifetable could be improved. We can define a general $_t p_x$ function to calculate the cumulative survival probabilities.


```{r}
tpxs_fn <- function(age, qx, x) {

  work_df=data.frame(age,qx)  
  qxs <- work_df[work_df$age >= x,]$qx
  Reduce("*",1-qxs,1,accumulate=TRUE)
  
}
p <- function(t){
  tpxs <- tpxs_fn(PMA92C10$age,PMA92C10$qx, 40)
  repeatLast(tpxs,t+1)[t+1]
#Note how this relies implicitly on the last qx being 1! If we were deploying this we shoud check and enforce this.  
  }

```

This doesn't change the results...

``` {r}
CF(5)
DCF(5)
DCF(10)
PV(100)
```

## Example - whole of life annuities 

We can easily enough adaptt this setup now to calculate whole of life annuities and/or deferred whole of life annuities

```{r}
C <- function(t) {1}

tpxs_fn <- function(age, qx, x, t) {

  work_df=data.frame(age,qx)  
  qxs <- work_df[work_df$age >= x,]$qx
  tpxs <- Reduce("*",1-qxs,1,accumulate=TRUE)
  repeatLast(tpxs,t+1)[t+1]
  #Note how this relies implicitly on the last qx being 1! If we were deploying this we shoud check and enforce this.  

  }

p <- function(t){
  tpxs_fn(PMA92C10$age,PMA92C10$qx, 40,t)
  }



v_generic <- function (t, i) {Reduce("*", 1/(1+(repeatLast(i,t+1))),1,accumulate = TRUE)[t+1]}

v <- function(t) {v_generic(t,(0.028))}


CF <- function(t) {(p(t)) * C(t)}
DCF <- function(t) {v(t) * p(t) * C(t)}

CF(5)
DCF(10)

PV <- function(n) {Reduce("+",Map(DCF,seq(0,n)),0)}

PV(100)



```


## dealing with different payment timings


For a start we might want to make a discount function that can deal with non-integer times. 


```{r}
interp_linear <- function(x, x0,y0, x1,y1)
  {
    y0 + (x-x0)/(x1-x0) * (y1-y0)
}
#Check this definition!
interp_geom <- function(x, x0,y0, x1,y1)
  {
  y0 * (y1/y0)^((x-x0)/(x1-x0))
}

interp_linear(1.25, 1,5, 2,10)
interp_linear(1.25, 1,1.05, 2,1.10)
interp_geom(1.25, 1,1.05, 2,1.10)
```

```{r}
interp_disc <- function(v,t) 
  {
    t0 <- floor(t)
    d0 <- repeatLast(v,t0+1)[t0+1]
    d1 <- repeatLast(v,t0+2)[t0+2]
    interp_geom(t, t0,d0,t0+1,d1)
  }

interp_disc(1+c(0.02, 0.03, 0.02, 0.04), 0.5)
interp_disc(1+c(0.02, 0.03, 0.02, 0.04), 1.25)


```

This should allow us to define the discount function $v$ over all positive reals, rather than just integers.


```{r}
v_generic <- function (t, i) {
  interp_disc(Reduce("*", 1/(1+i),1,accumulate = TRUE),t)
  }
v <- function(t) {v_generic(t,(0.028))}
v(1)
v(2)
v(1.5)
```


