---
title: "StochasticCashflows"
author: "DavidSalisbury"
date: "18 June 2016"
output: html_document
---

What if we want to consider the cashflows stochastically rather than using mean probability approach?

Let's consider the whole of life annuity example again...
```{r}
C <- function(t) {1}

tpxs_fn <- function(age, qx, x, t) {

  work_df=data.frame(age,qx)  
  qxs <- work_df[work_df$age >= x,]$qx
  tpxs <- Reduce("*",1-qxs,1,accumulate=TRUE)
  repeatLast(tpxs,t+1)[t+1]
  #Note how this relies implicitly on the last qx being 1! If we were deploying this we shoud check and enforce this.  

  }

tpxs_stoch_fn <- function(age, qx, x, t) {

  work_df=data.frame(age,qx)  
  qxs <- work_df[work_df$age >= x,]$qx
  tpxs <- Reduce("*",1-qxs,1,accumulate=TRUE)
  rf <- runif(1, min=0, max=1)
  stoch_tpxs <- ifelse(tpxs>=rf,1,0)
  repeatLast(stoch_tpxs,t+1)[t+1]
  #Note how this relies implicitly on the last qx being 1! If we were deploying this we shoud check and enforce this.  

  }

```

This is not very efficient - if we plan to use this in real life we would memoise, or otherwise make sure we are not recalculating the tpxs at each iteration.

```{r}

p <- function(t){
  tpxs_stoch_fn(PMA92C10$age,PMA92C10$qx, 40,t)
  }



v_generic <- function (t, i) {Reduce("*", 1/(1+(repeatLast(i,t+1))),1,accumulate = TRUE)[t+1]}

v <- function(t) {v_generic(t,(0.028))}


CF <- function(t) {(p(t)) * C(t)}
DCF <- function(t) {v(t) * p(t) * C(t)}

CF(5)
DCF(10)

PV <- function(n) {Reduce("+",Map(DCF,seq(0,n)),0)[[1]]}

PV(100)
PV(100)
PV(100)

```

Note how the three PV calculations for the annuity now have different results (unless your are very VERY lucky!)


We can now generate a sample of annuity values for a whole of life annuity discounted at 2.8% for a life aged 40 assuming mortality inline with PMA92 Calendar Year 2010.

WARNING - takes a little time to run.

```{r}

annuity_sample <- sapply(rep_len(100,100),PV)
annuity_sample[5]
mean(annuity_sample)
sd(annuity_sample)


```



```{r}
library(ggplot2)
ggplot() +
  geom_point(data = annuity_sample, aes(x = 1, y = annuity_sample))

```